---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# MiCARA
MiCARA (Microbiome Confounder-Adjusted Robust Association Analysis) provides an end-to-end framework for quality control, metadata imputation, differential abundance testing, cross-domain residual correlation analysis, and interactive Sankey network visualization.

<!-- badges: start -->
<!-- badges: end -->

## Recommended Workflow

MiCARA follows a streamlined 5-step pipeline:

1. **`validate_inputs()`** — check data formatting, reconstructs raw counts from relative abundances/proportions if needed and filters out high-missingness covariates (>30%).
2. **`impute_metadata()`** *(Optional)* — Imputes remaining missing metadata values (<30%).
   > **When to skip:** Skip if your metadata is 100% complete or if your study protocol requires Complete Case Analysis (CCA).
3. **`run_diff_abundance()`** — Performs ANCOM-BC2 differential abundance on taxa and pathways.
4. **`compute_residualCLR_correlations()**` — Infers residual cross-domain correlations while adjusting for disease covariates.
5. **`plot_interaction_sankey()**` — Renders interactive Sankey network diagrams for significant interactions.


## Installation

You can install the development version of MiCARA from GitHub with:

``` {r, eval = FALSE}
# Install devtools if not already installed
if (!requireNamespace("devtools", quietly = TRUE)) {
    install.packages("devtools")
}

# Install MiCARA from GitHub
devtools::install_github("BushnaqSafa/MiCARA")
```


### Quickstart Example

```{r, eval = FALSE}
library(MiCARA)

# 1. Validate & filter inputs
obj <- validate_inputs(
  taxa_mat = taxa_mat, 
  pathway_mat = pathway_mat, 
  metadata = metadata, 
  disease_col = "disease"
)

# 2. Impute missing metadata (optional)
obj <- impute_metadata(obj)

# 3. Perform differential abundance on taxa and pathways
diff_res <- run_diff_abundance(obj, feature_type = c("taxa", "pathways"))

# 4. Infer layer-2 residual cross-domain correlations
interactions <- compute_residualCLR_correlations(obj, diffab_obj = diff_res)

# 5. Visualize interactions in a Sankey network
plot_interaction_sankey(interactions, disease = "UC", direction = "both")

```



