---
title: "Getting-started with MiCARA"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting-started with MiCARA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(MiCARA)
```

## Pipeline Workflow & Decision Points

MiCARA is designed to be flexible depending on your dataset's completeness and clinical reporting requirements.

### Step 1: Input Validation

`validate_inputs()` ensures sample ID alignment across taxonomic matrices, pathway matrices, and sample metadata. It checks relative vs. absolute abundances, flags high-missingness covariates (>30%), and records retained confounders.

```{r validate-step, eval = TRUE}
# Load example dataset included in MiCARA
data("example_taxa", package = "MiCARA")
data("example_pathways", package = "MiCARA")
data("example_metadata", package = "MiCARA")

# Validate inputs and filter unusable covariates
micara_obj <- validate_inputs(
  taxa     = example_taxa,
  pathways = example_pathways,
  metadata = example_metadata,
  disease_col = "disease"
)
```

### Step 2: Handling Missing Metadata (Optional)

Covariate missingness can lead to sample loss during linear model fitting (listwise deletion). MiCARA provides impute_metadata() to address this via MICE:
**When to use impute_metadata():
Host covariates (e.g., age_category, BMI) have low-to-moderate missingness (<30%).
You want to preserve sample size and statistical power.

**When to skip impute_metadata():
Metadata contains zero missing values (NA).
Institutional protocols mandate Complete Case Analysis (CCA).


Note: If you skip imputation, downstream functions will check for NAs in requested covariates and drop or warn on problematic variables before model fitting.

```{r impute-step, eval = TRUE}
# Optional: Impute remaining missing metadata values (<30%)
micara_obj <- impute_metadata(micara_obj)
```

### Step 3: Differential Abundance Analysis

`run_diff_abundance()` runs bias-corrected log-linear models across taxa, pathways, or both simultaneously using ANCOM-BC2.

```{r diff-step, eval = TRUE}
# Run differential abundance on both feature types
diff_results <- run_diff_abundance(
  micara_obj      = micara_obj,
  feature_type    = c("taxa", "pathways"),
  reference_level = "healthy",
  use_robust      = TRUE
)

# Inspect significant taxa
head(diff_results$taxa$significant)

# Inspect significant pathways
head(diff_results$pathways$significant)
```

### Step 4: Confounder-Adjusted Interaction Inference
After identifying robust differentially abundant features, compute_residual_correlations() evaluates taxon–pathway interactions within each disease cohort:
1. Performs Centered Log-Ratio (CLR) transformations on full feature matrices.
2. Filters to disease-specific significant taxa and pathways.
3. Residualizes feature abundances against host covariates (e.g., age_category, BMI, gender, study_name) using QR decomposition.
4. Computes pairwise Spearman correlations on residuals and applies Benjamini-Hochberg (BH) adjustment within each cohort.

```{r interactions, eval = TRUE}
# Compute residual correlations per disease
interactions <- compute_residualCLR_correlations(
  micara_obj   = micara_obj,
  diffab_obj   = diff_results,
  covariates   = c("age_category", "BMI", "gender", "study_name"),
  q_cutoff     = 0.05,
  r_cutoff     = 0.20,
  verbose      = TRUE
)

# inspect available disease cohorts in the output list
names(interactions)
```

```{r inspect-diagnostics}

# View cohort-level analysis summary and covariate usage
attr(interactions, "summary")

# View analysis settings and CLR transformation diagnostics
attr(interactions, "settings")

# Find the first analyzed cohort dynamically
analyzed_cohorts <- names(interactions)[sapply(interactions, function(x) is.list(x) && x$status == "analysed")]

if (length(analyzed_cohorts) > 0) {
  first_cohort <- analyzed_cohorts[1]
  message("Displaying links for cohort: ", first_cohort)
  head(interactions[[first_cohort]]$links)
}
```
### Step 5: Visualize Taxa-Pathway Interactions

Use `plot_interaction_sankey()` to display interactive D3 networks connecting significant taxa to functional pathways:

```{r plot-sankey, eval=FALSE}
# Interactive visualization
plot_interaction_sankey(
  interactions      = interactions,
  disease           = names(interactions)[1], # Target disease cohort (e.g., "UC")
  direction         = "both",            # Options: "positive", "negative", or "both"
  top_n_per_pathway = 5,                 # Top N correlated taxa per pathway
  font_size         = 12,                # Text label size in pixels
  node_width        = 35                 # Width of Sankey block nodes
)

# Optional: Save interactive HTML and static PNG exports
# Create the results directory first
dir.create("results", showWarnings = FALSE, recursive = TRUE)

plot_interaction_sankey(
  interactions      = interactions,
  disease           = names(interactions)[1],
  save_path         = "results/sankey_network"
)
```



